---
title: "Data analysis exercise (Exercise 2.2)"
author: "Oscar Brück"
date: "02.02.2017"
output:
  html_document:
    theme: journal
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6
    fig_height: 4
    code_folding: hide
---

#Exercise 2 Data analysis

##Read the dataset from the local folder and analyze the structure of the dataset
```{r echo = TRUE}
setwd("/Users/obruck/Desktop/Open data course/IODS-project/data/")  
learning2014 <- read.csv("learning2014.csv", header=TRUE)  
str(learning2014)
```

```{r echo = TRUE}
dim(learning2014)  
```
  
The "learning2014" dataset constitutes of 166 observations and 8 variables. The variables (gender, Age, Attitude, Deep, Stra, Surf and Points) describe the results of a query study made as a collaboration international project with teachers.  

  + Gender = 1 corresponds to male and 2 to female students  
  + Age = the age of the students  
  + Attitude = Global attitude toward statistics, computed as a sum of numerous variables  
  + Deep = Deep approach, computed as a sum of numerous variables  
  + Stra = Surface approach, computed as a sum of numerous variables  
  + Surf = Surface approach, computed as a sum of numerous variables  
  + Points = Max points of social studies  

##Show a graphical overview of the data and show summaries of the variables in the data  
```{r echo = TRUE}
summary(learning2014)
```
  
There are 110 female and 56 male students. The median age is 22 years and range [17-55]. Students got a median of 44 points [19-59] in the deep approach (Deep), 25.50 [10-40] in the strategic approach (Stra), 34 points [19-52] in the surface approach (Surf). Their max points were 23 [7-33] as a median (Points).
```{r echo = TRUE}
pairs(learning2014[-1], col=learning2014$gender)
```
  
The data is visualised according to gender (males as black dots and females as red dots) on multiple bivariable correlation scatter plots.  

The data can also be visualised using the "GGally" and "ggplot2" packages
```{r echo = TRUE}
library("GGally")  
library("ggplot2")  
p <- ggpairs(learning2014[-1], mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))  
p
```
  
Now we can see that male students have a higher score in attitude questions and female students in strategic questions. We can also see that Variables Points and Attitude have the highest (cor: 0.43) and Surf and Deep second highest correlation factor (cor: 0.32).  


##Regression model  
The "Points" student achieve is tested with a multivariate linear regression model using Attitude, Stra and Surf as independent variables. The model tries to describe any linear correlation between the independent and dependent variables.
```{r echo = TRUE}
points_regression <- lm(Points ~ Attitude + Stra + Surf, data = learning2014)  
summary(points_regression)  
```  
In the first model with all three variables, Attitude explains the Points with a coefficient estimate of 0.34 (p<0.0001). Other variables Stra and Surf do not explain Points with a statistical significance. Thus, first Surf is removed and the model is tested again.
```{r echo = TRUE}
points_regression <- lm(Points ~ Attitude + Stra, data = learning2014)  
summary(points_regression)    
```  
Attitude is still the only relevant variable and, thus, Stra is removed and the model is run with Attitude alone (coefficient 0.35, p<0.0001).
```{r echo = TRUE}
points_regression <- lm(Points ~ Attitude, data = learning2014) 
summary(points_regression)      
```  
The adjusted R-squared score is 0.19 meaning that using this model, Attitude explains 19% of the total variance of the dependent variable Points.  

##Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage  
```{r echo = TRUE}
plot(points_regression, which = c(1,2,5))
```  

Residuals of the linear regression model describe the validity of the model. Residuals should be normally distributed, they should not correlate with each others, have a constant varaiance and their size should not depend on the size of the independent variable.  
The QQ-plot describes the distribution of the residuals. Our model is normally divided as the measurements are set on straight increasing line.  
The Residuals vs Fitted plot describes how the residuals depend on the explanatory variable. According to the plot, the resudials are resonably evenly distributed on both sides of the residual level = 0 line and, thus, do not depend on the size of the explanatory variable.  
The Residuals vs Leverage plot can help identify, which observations have an unusually high impact on the model. According to the plot, all measurements are divided between [0,0.05] and thus none have an unusual impact on the model.